Weakly Nonlinear MRI: Notes

### 2014-02-10 notes/mri_notes.tex
fixed lorentz force term

@@

\begin{equation}
\partial_t \mathbf{u} \, + \, \mathbf{u} \cdot \nabla \mathbf{u} \, + 2\Omega_0\left(v_r \mathbf{\hat{\phi}} - v_{\phi}\mathbf{\hat{r}}\right) \, = \, - r\left(\Omega^2 - \Omega_0^2\right)\mathbf{\hat{r}} \, -\frac{1}{\rho}\nabla P \, +\frac{1}{\rho_b}\nabla P_b \,  - \, \frac{1}{8 \pi \rho}\nabla B^2 + \frac{1}{4 \pi \rho} \left(\mathbf{B} \cdot \nabla\right)\mathbf{B}  \, + \, \nu\nabla^2 \mathbf{u} \,
\end{equation}

@@

And our momentum equation now reads: 
\[\partial_t \mathbf{u} \, + \, \mathbf{u} \cdot \nabla \mathbf{u} \, + 2\Omega_0\left(u_x \mathbf{\hat{y}} - u_{y}\mathbf{\hat{x}}\right) \, = \, 2\Omega_0^2 x q \mathbf{\hat{x}} \, -\frac{1}{\rho}\nabla P \, +\frac{1}{\rho_b}\nabla P_b \,  - \, \frac{1}{8 \pi \rho}\nabla B^2 + \frac{1}{4 \pi \rho} \left(\mathbf{B} \cdot \nabla\right)\mathbf{B}  \, + \, \nu\nabla^2 \mathbf{u} \,\]

@@

Where $v_A$ is the Alfven speed, which is $v_A = \frac{B_0^2}{\rho_0}$. \\

@@

\[\frac{\partial \mathbf{\widetilde{u}}}{\partial \widetilde{t}} \, + \, \mathbf{\widetilde{u}} \cdot \widetilde{\nabla} \mathbf{\widetilde{u}} \, + 2\Omega_0\left(\widetilde{u_x} \mathbf{\hat{y}} - \widetilde{u_{y}}\mathbf{\hat{x}}\right) \, = \, 2\Omega_0^2 \widetilde{x} q \mathbf{\hat{x}} \, -\frac{1}{\widetilde{\rho}}\widetilde{\nabla} \widetilde{P} \, +\frac{1}{\widetilde{\rho_b}}\widetilde{\nabla} \widetilde{P_b} \,  - \, \frac{1}{8 \pi \widetilde{\rho}}\widetilde{\nabla} \widetilde{B}^2 + \frac{1}{4 \pi \widetilde{\rho}} \left(\mathbf{\widetilde{B}} \cdot \widetilde{\nabla}\right)\mathbf{\widetilde{B}}  \, + \, \nu \widetilde{\nabla}^2 \mathbf{\widetilde{u}} \,\]

@@

%\begin{equation}
$\left(\Omega_0^2 r_0 \delta\right) \partial_t \mathbf{u} \, + \, \left(\Omega_0^2 r_0 \delta\right)\mathbf{u} \cdot \nabla \mathbf{u} \,+\, \left(\Omega_0^2 r_0 \delta\right) 2 u_x \mathbf{\hat{y}} \,-\, \left(\Omega_0^2 r_0 \delta\right) 2 u_y \mathbf{\hat{x}} \, = \, \left(\Omega_0^2 r_0 \delta\right) 2 x q \mathbf{\hat{x}} \,-\, \Omega_0^2 r_0 \left(\frac{\epsilon^2}{\delta}\right)\frac{1}{\rho} \nabla P  \,+\, \Omega_0^2 r_0 \left(\frac{\epsilon^2}{\delta}\right)\frac{1}{\rho_b} \nabla P_b \, - \, \frac{1}{8\pi}\Omega_0^2 r_0 \left(\frac{\epsilon^2}{\delta}\right) \frac{1}{\beta} \frac{1}{\rho} \nabla B^2 \, + \, \frac{1}{4\pi} \Omega_0^2 r_0 \left(\frac{\epsilon^2}{\delta}\right) \frac{1}{\beta} \frac{1}{\rho} \left(\mathbf{B} \cdot \nabla \right)\mathbf{B} \, + \, \nu\left(\frac{\Omega_0}{r_0 \delta}\right) \nabla^2 \mathbf{u}$ \\
%\end{equation}

@@

$\partial_t \mathbf{u} \, + \, \mathbf{u} \cdot \nabla \mathbf{u} \, + \, 2 u_x \mathbf{\hat{y}} - 2 u_y \mathbf{\hat{x}} \, = \, 2 x q \mathbf{\hat{x}} \, - \, \frac{\epsilon^2}{\delta^2} \frac{1}{\rho} \nabla P \, + \,  \frac{\epsilon^2}{\delta^2} \frac{1}{\rho_b} \nabla P_b \, - \, \frac{1}{8\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \nabla B^2 \, + \, \frac{1}{4\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \left(\mathbf{B} \cdot \nabla \right)\mathbf{B} \, + \, \frac{1}{R_e} \nabla^2 \mathbf{u}$ \\

@@

$\partial_t u_x \, + \, \left(\mathbf{u} \cdot \nabla\right)u_x \, - \, 2 u_y \, = \, 2 x q \, - \, \frac{\epsilon^2}{\delta^2} \frac{1}{\rho} \partial_x P \, + \, \frac{\epsilon^2}{\delta^2} \frac{1}{\rho_b} \partial_x P_b \, -\, \frac{1}{8\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \partial_x \left(\mathbf{B}^2\right) \, + \, \frac{1}{4\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \left(\mathbf{B} \cdot \nabla\right) B_x \, + \, \frac{1}{R_e}\nabla^2 u_x  $ \\

$\partial_t u_y \, + \, \left(\mathbf{u} \cdot \nabla\right)u_y \, + \, 2u_x =  \, - \, \frac{\epsilon^2}{\delta^2} \frac{1}{\rho} \partial_y P \, -\, \frac{1}{8\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \partial_y \left(\mathbf{B}^2\right) \, + \, \frac{1}{4\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \left(\mathbf{B} \cdot \nabla\right) B_y \, + \, \frac{1}{R_e}\nabla^2 u_y  $ \\

$\partial_t u_z \, + \, \left(\mathbf{u} \cdot \nabla\right)u_z \,  =  \, - \, \frac{\epsilon^2}{\delta^2} \frac{1}{\rho} \partial_z P \, -\, \frac{1}{8\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \partial_z \left(\mathbf{B}^2\right) \, + \, \frac{1}{4\pi} \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{\rho} \left(\mathbf{B} \cdot \nabla\right) B_z \, + \, \frac{1}{R_e}\nabla^2 u_z $ \\

@@

Similarly, the magnetic tension term expands to become: \\

$\left(\mathbf{B} \cdot \nabla \right)\mathbf{B} \, = \, \left(\left(B_0 \mathbf{\hat{z}} + \mathbf{B_1} \right)\cdot \nabla \right)\left(B_0 \mathbf{\hat{z}} + \mathbf{B_1} \right) \, = \, B_0 \partial_z \mathbf{B_1} \, + \, \mathbf{B_1} \cdot \nabla \mathbf{B_1}$ \\

Because the $\nabla B_0 \mathbf{\hat{z}}$ terms are 0.

The magnetic pressure term expands to become: \\

$\nabla \left(B_0 \mathbf{\hat{z}} + \mathbf{B_1}\right)^2 $ \\

>>>>

However, this formulation allows the magnetic pressure term to have a component parallel to the magnetic field. Clearly $\mathbf{J} \times \mathbf{B}$ is always perpendicular to the field. This will become problematic if we leave these two terms separated \-- terms will cancel in the perturbation expansion that should not. So for the time being, we retain the Lorentz force as $\frac{1}{4 \pi \rho} \left( \nabla \times \mathbf{B} \right) \times \mathbf{B}$. \\

@@

\begin{equation}
\partial_t \mathbf{u} \, + \, \mathbf{u} \cdot \nabla \mathbf{u} \, + 2\Omega_0\left(v_r \mathbf{\hat{\phi}} - v_{\phi}\mathbf{\hat{r}}\right) \, = \, - r\left(\Omega^2 - \Omega_0^2\right)\mathbf{\hat{r}} \, -\frac{1}{\rho}\nabla P \, +\frac{1}{\rho_b}\nabla P_b \,  + \, \frac{1}{4 \pi \rho} \left( \nabla \times \mathbf{B} \right) \times \mathbf{B}  \, + \, \nu\nabla^2 \mathbf{u} \,
\end{equation}

@@

And our momentum equation now reads: 
\[\partial_t \mathbf{u} \, + \, \mathbf{u} \cdot \nabla \mathbf{u} \, + 2\Omega_0\left(u_x \mathbf{\hat{y}} - u_{y}\mathbf{\hat{x}}\right) \, = \, 2\Omega_0^2 x q \mathbf{\hat{x}} \, -\frac{1}{\rho}\nabla P \, +\frac{1}{\rho_b}\nabla P_b \,  + \, \frac{1}{4 \pi \rho} \left( \nabla \times \mathbf{B} \right) \times \mathbf{B} + \, \nu\nabla^2 \mathbf{u} \,\]

@@

Where $v_A$ is the Alfven speed, which is ${v_A}^2 = \frac{B_0^2}{\rho_0}$. \\

@@

\[\frac{\partial \mathbf{\widetilde{u}}}{\partial \widetilde{t}} \, + \, \mathbf{\widetilde{u}} \cdot \widetilde{\nabla} \mathbf{\widetilde{u}} \, + 2\Omega_0\left(\widetilde{u_x} \mathbf{\hat{y}} - \widetilde{u_{y}}\mathbf{\hat{x}}\right) \, = \, 2\Omega_0^2 \widetilde{x} q \mathbf{\hat{x}} \, -\frac{1}{\widetilde{\rho}}\widetilde{\nabla} \widetilde{P} \, +\frac{1}{\widetilde{\rho_b}}\widetilde{\nabla} \widetilde{P_b} \,  + \, \frac{1}{4 \pi \widetilde{\rho}} \left( \widetilde{\nabla} \times \mathbf{\widetilde{B}} \right) \times \mathbf{\widetilde{B}}  \, + \, \nu \widetilde{\nabla}^2 \mathbf{\widetilde{u}} \,\]

@@

%\begin{equation}
$\left(\Omega_0^2 r_0 \delta\right) \partial_t \mathbf{u} \, + \, \left(\Omega_0^2 r_0 \delta\right)\mathbf{u} \cdot \nabla \mathbf{u} \,+\, \left(\Omega_0^2 r_0 \delta\right) 2 u_x \mathbf{\hat{y}} \,-\, \left(\Omega_0^2 r_0 \delta\right) 2 u_y \mathbf{\hat{x}} \, = \, \left(\Omega_0^2 r_0 \delta\right) 2 x q \mathbf{\hat{x}} \,-\, \Omega_0^2 r_0 \left(\frac{\epsilon^2}{\delta}\right)\frac{1}{\rho} \nabla P  \,+\, \Omega_0^2 r_0 \left(\frac{\epsilon^2}{\delta}\right)\frac{1}{\rho_b} \nabla P_b \, - \, \frac{\epsilon^2 \Omega_0^2 r_0}{\beta \delta} \frac{1}{4 \pi \rho} \left( \nabla \times \mathbf{B} \right) \times \mathbf{B} \, + \, \nu\left(\frac{\Omega_0}{r_0 \delta}\right) \nabla^2 \mathbf{u}$ \\
%\end{equation}

@@

$\partial_t \mathbf{u} \, + \, \mathbf{u} \cdot \nabla \mathbf{u} \, + \, 2 u_x \mathbf{\hat{y}} - 2 u_y \mathbf{\hat{x}} \, = \, 2 x q \mathbf{\hat{x}} \, - \, \frac{\epsilon^2}{\delta^2} \frac{1}{\rho} \nabla P \, + \,  \frac{\epsilon^2}{\delta^2} \frac{1}{\rho_b} \nabla P_b \, - \, \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{4 \pi \rho} \left( \nabla \times \mathbf{B} \right) \times \mathbf{B} \, + \, \frac{1}{R_e} \nabla^2 \mathbf{u}$ \\

@@

$\partial_t u_x \, + \, \left(\mathbf{u} \cdot \nabla\right)u_x \, - \, 2 u_y \, = \, 2 x q \, - \, \frac{\epsilon^2}{\delta^2} \frac{1}{\rho} \partial_x P \, + \, \frac{\epsilon^2}{\delta^2} \frac{1}{\rho_b} \partial_x P_b \, +\, \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{4 \pi \rho} \left( \nabla \times \mathbf{B} \right) \times \mathbf{B} + \, \frac{1}{R_e}\nabla^2 u_x  $ \\

$\partial_t u_y \, + \, \left(\mathbf{u} \cdot \nabla\right)u_y \, + \, 2u_x =  \, - \, \frac{\epsilon^2}{\delta^2} \frac{1}{\rho} \partial_y P \, +\, \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{4 \pi \rho} \left( \nabla \times \mathbf{B} \right) \times \mathbf{B} \, + \, \frac{1}{R_e}\nabla^2 u_y  $ \\

$\partial_t u_z \, + \, \left(\mathbf{u} \cdot \nabla\right)u_z \,  =  \, - \, \frac{\epsilon^2}{\delta^2} \frac{1}{\rho} \partial_z P \, +\, \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{4 \pi \rho} \left( \nabla \times \mathbf{B} \right) \times \mathbf{B} \, + \, \frac{1}{R_e}\nabla^2 u_z $ \\

@@

%Similarly, the magnetic tension term expands to become: \\

%$\left(\mathbf{B} \cdot \nabla \right)\mathbf{B} \, = \, \left(\left(B_0 \mathbf{\hat{z}} + \mathbf{B_1} \right)\cdot \nabla \right)\left(B_0 \mathbf{\hat{z}} + \mathbf{B_1} \right) \, = \, B_0 \partial_z \mathbf{B_1} \, + \, \mathbf{B_1} \cdot \nabla \mathbf{B_1}$ \\

%Because the $\nabla B_0 \mathbf{\hat{z}}$ terms are 0.

%The magnetic pressure term expands to become: \\

%$\nabla \left(B_0 \mathbf{\hat{z}} + \mathbf{B_1}\right)^2 $ \\

The Lorentz force term $\frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{4 \pi \rho} \left( \nabla \times \mathbf{B} \right) \times \mathbf{B}$ expands as follows: \\

$ \left( \nabla \times \left(B_0 \mathbf{\hat{z}} + \mathbf{B_1}\right) \right) \times \left(B_0 \mathbf{\hat{z}} + \mathbf{B_1}\right) \, = \, \left( \nabla \times B_0 \mathbf{\hat{z}} + \nabla \times \mathbf{B_1} \right) \times \left(B_0 \mathbf{\hat{z}} + \mathbf{B_1}\right) \, $ \\

$= \, \left( \nabla \times B_0 \mathbf{\hat{z}} + \nabla \times \mathbf{B_1} \right) \times B_0 \mathbf{\hat{z}} + \left( \nabla \times B_0 \mathbf{\hat{z}} + \nabla \times \mathbf{B_1} \right) \times \mathbf{B_1} \, $ \\

$= \, \nabla \times B_0 \mathbf{\hat{z}} \times B_0 \mathbf{\hat{z}} \, +\, \nabla \times \mathbf{B_1} \times B_0 \mathbf{\hat{z}} \, + \, \nabla \times B_0 \mathbf{\hat{z}} \times \mathbf{B_1} \, + \, \nabla \times \mathbf{B_1} \times \mathbf{B_1}$\\

$= \, -\frac{\nabla B_0^2}{2} \, + \, B_0 \mathbf{\hat{z}} \cdot \nabla \left(B_0 \mathbf{\hat{z}}\right) \, - \, \nabla\left(B_0 \mathbf{\hat{z}} \cdot \mathbf{B_1}\right) \, + \, B_0 \mathbf{\hat{z}} \cdot \nabla \mathbf{B_1} \, + \, \mathbf{B_1} \cdot \nabla B_0 \mathbf{\hat{z}} \, + \, \mathbf{B_1} \times \nabla \times B_0 \mathbf{\hat{z}} \, + \, \nabla \times B_0 \mathbf{\hat{z}} \times \mathbf{B_1} \, - \, \frac{\nabla \mathbf{B_1}^2}{2} \, + \, \mathbf{B_1} \cdot \nabla \mathbf{B_1}$ \\

Now the $\nabla B_0 \mathbf{\hat{z}}$ terms are zero, and $\mathbf{B_1} \times \nabla \times B_0 \mathbf{\hat{z}} \, = \, - \nabla \times B_0 \mathbf{\hat{z}} \times \mathbf{B_1}$, so those two terms cancel one another. So we have: \\

$ = \, \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{4 \pi \rho} \left(-\frac{\nabla B_0^2}{2} \, - \, \nabla \left(B_0 \mathbf{\hat{z}}\cdot \mathbf{B_1}\right)\, + \, B_0 \mathbf{\hat{z}} \cdot \nabla \mathbf{B_1} \, - \, \frac{\mathbf{B_1}^2}{2} \, + \, \mathbf{B_1}\cdot \nabla \mathbf{B_1} \, \right)$ \\

And finally, the Lorentz force term becomes: \\

$\frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{4 \pi \rho} \left( \nabla \times \mathbf{B} \right) \times \mathbf{B} \, = \, \frac{\epsilon^2}{\delta^2} \frac{1}{\beta} \frac{1}{4 \pi \rho} \left(-\frac{\nabla B_0^2}{2} \,  - \, \frac{\mathbf{B_1}^2}{2} \, + \, \mathbf{B_1}\cdot \nabla \mathbf{B_1} \, \right)$ \\
